Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-07-26 paper.tex
Updates in Multiple read model

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{multirow}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available RAM (in B).
\item $m'$: amount of used RAM (in B). $m'\leq m$.
\end{itemize}

The following relations are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, we assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}=\nu$.
\end{itemize}
%\todo{Make a figure to illustrate the notations.}

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), complete block rows (case
2) or complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. However, such cases are a bit pathological and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}[b]
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/incomplete-rows.pdf_tex}
\hfill
\def\svgwidth{0.3\columnwidth}
\input{figures/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}
To summarize, the amount of memory used $m'$ is set as follows in each of the 3 cases:
%% \begin{equation*}
%%   m_{case 1}' = bd^3\floor*{\frac{m}{d^3b}}; m_{case 2}' = bd^3\sqrt[3]{n}\floor*{\frac{m}{d^3b\sqrt[3]{n}}}; m_{case 3}' = bd^3\sqrt[3]{n}^2\floor*{\frac{m}{d^3b\sqrt[3]{n}^2}}
%% \end{equation*}
%% which gives:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}

@@

Finally, the total number of seeks performed by cluster reads to reconstruct the image is:
%% \begin{equation*}
%% N_\mathrm{cr} =
%% \begin{cases}
%%   n + \sqrt[3]{n}^2 \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}d^2 & \text{if } \frac{m}{d^3b} < \sqrt[3]{n}\\
%%   n + \sqrt[3]{n} \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}d      & \text{if } \sqrt[3]{n} \leq \frac{m}{d^3b} < \sqrt[3]{n}^2\\
%%   n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{m}{d^3b} < n\\
%%   n + 1                                                          & \text{otherwise}
%% \end{cases} \label{eq:seeks-cluster-reads}
%% \end{equation*}
%% And since $d=\frac{D}{\sqrt[3]{n}}$ and $R=D^3$, we finally have:
\begin{equation}
N_\mathrm{cr} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } \frac{mn}{Rb} < \sqrt[3]{n}\\
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \sqrt[3]{n} \leq \frac{mn}{Rb} < \sqrt[3]{n}^2\\
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{mn}{Rb} < n\\
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}

@@

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads.}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_split\_name\_list = sort blocks by increasing (z,y,x) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

@@

>>>>

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{stmaryrd} % for llbracket and rrbracket
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available memory (in B).
\item $m'$: amount of used memory (in B), $m'\leq m$.
\end{itemize}

For the sake of mathematical simplicity, we assume that
the reconstructed image and blocks are cubic. Under this assumption,
we also have the following relations:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}$.
\end{itemize}
%\todo{Make a figure to illustrate the notations.}

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), complete block rows (case
2) or complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. However, such cases are a bit pathological and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/incomplete-rows.pdf_tex}
\hfill
\def\svgwidth{0.3\columnwidth}
\input{figures/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}
To summarize, the amount of memory used $m'$ is set as follows in each of the 3 cases:
%% \begin{equation*}
%%   m_{case 1}' = bd^3\floor*{\frac{m}{d^3b}}; m_{case 2}' = bd^3\sqrt[3]{n}\floor*{\frac{m}{d^3b\sqrt[3]{n}}}; m_{case 3}' = bd^3\sqrt[3]{n}^2\floor*{\frac{m}{d^3b\sqrt[3]{n}^2}}
%% \end{equation*}
%% which gives:
\begin{equation*}
  m_1' = \frac{Rb}{n}\floor*{\frac{mn}{Rb}};
  m_2' = \frac{Rb}{\sqrt[3]{n}^2}\floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \frac{Rb}{\sqrt[3]{n}}\floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}

@@

Finally, the total number of seeks performed by cluster reads to reconstruct the image is:
%% \begin{equation*}
%% N_\mathrm{cr} =
%% \begin{cases}
%%   n + \sqrt[3]{n}^2 \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}d^2 & \text{if } \frac{m}{d^3b} < \sqrt[3]{n}\\
%%   n + \sqrt[3]{n} \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}d      & \text{if } \sqrt[3]{n} \leq \frac{m}{d^3b} < \sqrt[3]{n}^2\\
%%   n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{m}{d^3b} < n\\
%%   n + 1                                                          & \text{otherwise}
%% \end{cases} \label{eq:seeks-cluster-reads}
%% \end{equation*}
%% And since $d=\frac{D}{\sqrt[3]{n}}$ and $R=D^3$, we finally have:
\begin{equation}
N_\mathrm{cr} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } \frac{mn}{Rb} < \sqrt[3]{n}\\
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \sqrt[3]{n} \leq \frac{mn}{Rb} < \sqrt[3]{n}^2\\
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{mn}{Rb} < n\\
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}
\todo{Is $N_{cr}$ a continuous function of m?}

@@

\begin{algorithm}[h]
  \caption{Buffered merging of blocks with multiple reads}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_split\_name\_list = sort blocks by increasing (z,y,x) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split
        \STATE extract all indexes in next write range
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

For simplicity, we assume that $m'$ represents an integer number $k$
of block rows (case 1, $k<\sqrt[3]{n}$), of complete rows (case 2,
$k<d$), of tile rows (case 3, $k<\sqrt[3]{n}$), of slices (case 4,
$k<d$) or of block slices (case 5, $k<\sqrt[3]{n}$) \todo{make a
  figure}. In each of these 5 cases, we define $v$ as follows:
\begin{equation*}
  v_1=db \text{ ; }  v_2=Db \text{ ; } v_3=Ddb \text{ ; } v_4=D^2b \text{ ; } v_5=D^2db
\end{equation*}
so that we have:
\begin{equation*}
k_i=\floor*{\frac{m}{v_i}} \quad \mathrm{and} \quad m_i'=k_iv_i, \quad i \in \llbracket 1,5 \rrbracket
\end{equation*}
Using these notations, the number of seeks performed by multiple reads is:
\begin{equation*}
  N_\mathrm{MR} =
\begin{cases}
  \floor*{\frac{Rb}{m_1'}}(k_1+1)+1
  + n \text{ mod } k_1& \text{if } d \leq \frac{m}{b} < D\\
  \ceil*{\frac{Rb}{m_2'}}(\sqrt[3]{n}+1)   & \text{if } D \leq \frac{m}{b} < Dd\\
  \floor*{\frac{Rb}{m_3'}}(k_3\sqrt[3]{n}+1)+1+\\
  \quad \quad \sqrt[3]{n}\left(\frac{n}{\sqrt[3]{n}} \text{ mod } k_3\right)&\
  \text{if } Dd \leq \frac{m}{b} < D^2\\
  \ceil*{\frac{Rb}{m_4'}}(\sqrt[3]{n}^2+1) & \text{if } D^2 \leq \frac{m}{b} < D^2d\\
  \floor*{\frac{Rb}{m_5'}}(k_5\sqrt[3]{n}^2+1)+1+&\\
  \quad \quad \sqrt[3]{n}^2\left(\frac{n}{\sqrt[3]{n}^2} \text{ mod } k_5\right)&\
  \text{if } D^2d \leq \frac{m}{b} < R
\end{cases} 
\end{equation*}

which gives, using $R$ and $n$ as main variables:
\begin{equation}
N_\mathrm{MR} =
\begin{cases}
  \floor*{\frac{Rb}{m_1'}}\left(\frac{m_1'\sqrt[3]{n}}{\sqrt[3]{R}b}+1\right)+1+\\
  \quad \quad \quad n \text{ mod }\floor*{\frac{m\sqrt[3]{n}}{\sqrt[3]{R}b}}\
  & \text{if } \sqrt[3]{\frac{R}{n}} \leq \frac{m}{b} < \sqrt[3]{R}\\

  \ceil*{\frac{Rb}{m_2'}}\left(\sqrt[3]{n}+1\right)\
  & \text{if } \sqrt[3]{R} \leq \frac{m}{b} < \frac{\sqrt[3]{R}^2}{\sqrt[3]{n}}\\

  \floor*{\frac{Rb}{m_3'}}\left(\frac{m'_3\sqrt[3]{n}^2}{\sqrt[3]{R}^2b}+1\right)+1+\\
  \sqrt[3]{n}\left( \frac{n}{\sqrt[3]{n}} \text{ mod } \floor*{\frac{m\sqrt[3]{n}}{\sqrt[3]{R}^2b}} \right)
  & \text{if } \frac{\sqrt[3]{R}^2}{\sqrt[3]{n}} \leq \frac{m}{b} < \sqrt[3]{R}^2\\

  \ceil*{\frac{Rb}{m_4'}}\left(\sqrt[3]{n}^2+1\right)\
  & \text{if } \sqrt[3]{R}^2 \leq \frac{m}{b} < \frac{R}{\sqrt[3]{n}}\\

  \ceil*{\frac{Rb}{m_5'}}\left(\frac{m_5'n}{Rb}+1\right)+1+\\
  \sqrt[3]{n}^2\left( \frac{n}{\sqrt[3]{n}^2} \text{ mod } \floor*{\frac{m\sqrt[3]{n}}{Rb}}\right)
  & \text{if } \frac{R}{\sqrt[3]{n}} \leq \frac{m}{b} < R
\end{cases} \label{eq:seeks-multiple-reads}
\end{equation}

@@

\subsubsection{Simulations}
After running a few simulations, we saw that the comparison between
Cluster reads and Multiple reads essentially depends on the three
Cluster reads configurations.  When Cluster reads merge incomplete
block rows (case 1), they are always outperformed by Multiple reads.
When Cluster reads merge complete block rows (case 2), the comparison
between Cluster reads and Multiple reads depends on the values of $D$
and $n$. Figure~\ref{fig:model-comparison} shows two situations where
Cluster reads and Multiple reads are alternatively better than each
other. When Cluster reads merge complete block slices (case 3),
Cluster reads and Multiple reads are equivalent.

\begin{figure}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain.pdf}
  \includegraphics[width=0.45\columnwidth]{figures/model-big-brain-rescan.pdf}
  \caption{Number of seeks for Cluster reads vs Multiple reads when
    Cluster reads merge complete block rows (case 2), for D=3000 and
    b=2. Left: n=125; Right: n=64,000.}
  % Gnuplot file: scripts/model/model.gnplt
  \label{fig:model-comparison}
\end{figure}

\subsubsection{Experiments}
